---
title: "RLErestricted"
author: "José R. Ferrer-Paris"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RLErestricted}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RLErestricted)
```

## Introduction
In this vignette, we provide an example of using the tools within `RLErestricted` to
assess the spatial criteria of the [IUCN Red List of
Ecosystems](https://iucnrle.org/). These criteria assessproperties of the geographic
distribution size and its exposure to threats (Criterion B). This criterion requires the use of a map
of ecosystem distributions and qualitative data on threats.

This package relies on function from a couple of existing R packages. Particularly, the spatial data in vector format is read using the `sf` (simple features) package, and we use `units` to express the area calculations in standard units. Our functions make use of functions from the package `dplyr` that is part of the `tidyverse` collection of packages. 

```{r Loading packages, message=FALSE}
library(sf)
library(units)
library(dplyr)
```

## Importing data

We first import the spatial data we want to analyse. In this vignette, we 
will be using one example included with the package. 
We provide a simple dataset in `sf` format that represents the distribution of tropical glacier in the volcanic region between Peru and Chile. The data comes from the [Randolph Glacier Inventory, version 6](). 

```{r Loading our example distributions}
data("glaciers_on_volcanos")
glaciers_on_volcanos
```

## Assessing Criterion B (distribution size)

Criterion B utilizes measures of the geographic distribution of an ecosystem 
type to identify ecosystems that are at risk from catastrophic disturbances. 
This is done using two standardized metrics: the extent of occurrence (EOO) and 
the area of occupancy (AOO) [(Gaston and Fuller, (2009)](https://onlinelibrary.wiley.com/doi/10.1111/j.1365-2664.2008.01596.x/full), 
[Keith et al., (2013)](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0062111)). 

It must be emphasised that EOO and AOO are not used to estimate the mapped area
of an ecosystem like the methods we used in Criterion A; they are simply spatial
metrics that allow us to standardise an estimate of risk due to spatially explicit 
catastrophes [(Murray et al., (2017)](https://onlinelibrary.wiley.com/doi/10.1111/ddi.12533/abstract). 

Thus, it is critical that these measures are used consistently across all assessments, and
the use of non-standard measures invalidates comparison against the thresholds. 
Please refer to the guidelines ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf))
for more information on AOO and EOO.

### Subcriterion B1 (calculating EOO)

For subcriterion B1, we will need to calculate the extent of occurrence (EOO) of our
data. We begin by creating the minimum convex polygon enclosing all occurrences
of our focal ecosystem.

We can then calculate the area of the EOO polygon.


The calculated EOO area for this subset of mangroves in Victoria is ...
km2. This result can then be combined with additional information required
under B1(a-c) to assess the ecosystem under subcriteria B1.

### Subcriterion B2 (calculating AOO)

For subcriterion B2, we will need to calculate the number of 10x10 km grid cells
occupied by our distribution. We begin by creating the appopriate grid cells.

```{r Creating AOO grid, fig.width=7, fig.height=7}
AOO_grid <- create_AOO_grid(glaciers_on_volcanos)
print(AOO_grid)
```

Finally, we can use the created grid to calculate the AOO for our mangroves.

```{r Getting number of AOO grids}
AOO_summary(AOO_grid)
```


#### Grid uncertainty functions
Although we have a number for AOO , this may not be the actual minimum
AOO for this ecosystem, because the placement of the AOO grid will influence this
number. To take this into account, we have included a few additional functions.
We will be using `gridUncertainty` in this vignette.

`gridUncertainty` simply moves the AOO grid systematically (with a
small random movement around fixed points), and continues searching  for a minimum 
AOO until additional shifts no longer produce improved results. 

The result here shows that the minimum AOO does indeed change in response to 
where the AOO grid is placed, and that our original AOO value of  is
not the true minimum AOO. The true minimum AOO (and the one which should be used
for the assessment) is ....

It is also possible to plot out the grid which was used to produce the minimum 
AOO result, which can be used to report our final results.



#### Exporting results
The results of the above plot from `gridUncertainty` is a shapefile, which can
be exported with the `raster::shapefile` function.

#### One percent rule
In addition to the size of the grids used (which will be different for species 
assessments), there is also an option in the `makeAOOGrid` function to specify 
whether a minimum of percent of the grid cell area must be occupied before they 
are included as an AOO grid. Typically, this is only used in very special cases
for ecosystems with highly skewed distribution of patch sizes ([Bland et al., 2016](https://portals.iucn.org/library/sites/library/files/documents/2016-010.pdf)).

Here, we demonstrate the differences between including, or not including the one
percent rule:



There is an additional parameter - `percent` - which adjusts the threshold for
the AOO grids. Here, we set it to 0.1% to demonstrate its functionalities.



## 5. Final words
We have demonstrated in this vignette a typical workflow for assessing an
example ecosystem under the RLE. The methods outlined here can easily be adapted
for the Red List of Threatened Species as well, with slight adjustments in the
parameters (specifically the grid.size parameter for `getAOO` or
`gridUncertainty`). We hope that with this package, we help ensure a consistent
implementation for both red lists criteria, minimising any misclassification as
a result of using different software and methods.
